White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2015-09-06 papers/wdmerger_I/paper.tex

%==========================================================================
% Title
%==========================================================================
\title{Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification}

@@

For the energy equation, the central challenge is to write down a form of the 
discretized energy equation that explicitly conserves total energy when 
coupled to gravity. When gravity is included, the total energy is
\begin{equation}
  \rho E_{\text{tot}} = \rho E + \frac{1}{2}\rho\Phi = \frac{1}{2}\rho \mathbf{u}^2 + \rho e + \frac{1}{2}\rho\Phi, \label{eq:total_energy_gravity}
\end{equation}
where $\rho E$ is the total gas energy from the pure hydrodynamics equation,
and the factor of $1/2$ in the gravitational energy term is necessary
for simulations with self-gravity to prevent double-counting of
interactions.

Historically many simulation codes with gravity have not used a conservative formulation 
of the energy equation, but it is straightforward to do so. In particular, conservation of 
total energy requires only that a change in gravitational energy is compensated for by a 
change in gas energy, and that energy changes due to mass transfer are explicitly and 
exactly tracked. At each zone face in a given timestep, there is some transfer of mass $\Delta \rho \Delta V$ between 
the zones adjacent to that face, which is governed by the continuity equation. Suppose the 
transfer of mass is positive to the right, and the face is located at the (one-dimensional)
index $i-1/2$. Energy is transferred because an amount of mass $\Delta \rho_{i-1/2} \Delta V$ moves from 
potential $\Phi_{i-1}$ to potential $\Phi_{i}$; if these potentials are different, it requires some 
work exerted by or on the zone to push the mass to this different potential energy. Therefore 
the net change of gravitational energy is $\left[(1/2)\Delta \rho_{i-1/2} (\Phi_{i} - \Phi_{i-1})\right]$,
and if total energy is conserved, then the zone must have a change in gas energy 
$(\rho E)$ that is equal in magnitude and opposite in sign to the change in potential energy. 
Thus the update in a zone for the total energy that conserves $(\rho E_{\text{tot}})$ is:
\begin{equation}
  \Delta (\rho E) = -\frac{1}{2}\sum_{f} \Delta \rho_{f} (\Phi_{f+1/2} - \Phi_{f-1/2}),\label{eq:grav_energy_conservation_update}
\end{equation}
where the sum is over the cell faces $f$ and the indices $f+1/2$ and $f-1/2$ refer to 
the zone centers immediately to the left and right in the direction perpendicular to the face.
Note that while total energy is conserved, the gas energy is not because \autoref{eq:grav_energy_conservation_update} does not have the 
telescoping property of a flux formalism where energy gained by one zone is lost by another. 
The energy is being exchanged with a potential field that responds indirectly when the mass moves.

The change in the density $\Delta \rho_{f}$ is proportional to the mass flux through the surface with area $\Delta A$:
\begin{equation}
  \Delta \rho_f = \frac{\Delta t}{\Delta V} (\Delta A\, \rho v_{f}),\label{eq:mass_transfer}
\end{equation}
So as long as we record these fluxes through the zone faces after coming out of the hydrodynamics step, 
we are able to conserve the total energy completely (except for any energy loss or gain through 
physical domain boundaries). In order for the method to be second-order accurate in time, 
no change is needed to the fluxes since those are already time-centered 
in the hydrodynamics update, so we need only to use a time-centered $\Phi$ (which can be computed 
by averaging the time-level $n$ and $n+1$ potentials; we already have the latter because \castro\ 
re-computes the potential at the new time after the hydrodynamics step, and we can apply this energy 
at the end of the timestep). Using this conservative energy formulation is very important for simulations where the 
potential can change substantially over the course of the evolution, as massive amounts of energy can be lost. 
The standard approach is therefore unacceptable in the case of a problem like white dwarf mergers, and 
the fix to this energy leaking---evaluating the energy transfer at the six zone faces instead of 
the single zone center---has a very minor cost in terms of code complexity and computational time.

As observed by \cite{arepo}, when examining \autoref{eq:mass_transfer} one can see the relationship 
between this method and the more common (non-conservative) approach of evaluating the change in 
gas energy using the work done $(\mathbf{v} \cdot \rho \mathbf{g})$ by the gravitational force at the 
cell center. Analytically this form expresses the same core idea as \autoref{eq:grav_energy_conservation_update} 
via the work-energy theorem, but a major flaw is that it evaluates the energy change at the cell center 
when in fact the mass transfer is happening at the cell edges. However, there is a more fundamental issue. 
As observed by \citet[Chapter 4]{shu:1992} and \citet{marcello:2012}, the energy source term $(-\rho \mathbf{u}\cdot\nabla\Phi)$ 
includes a contribution that is not localizable and reflects the global nature of the gravitational potential. 
It is only by integrating this source term over all space that the correct gravitational potential energy 
for the system emerges. Integrating over only one zone cannot yield the correct description of this global 
energy flow, and indeed therefore can never obtain the factor of 1/2 that arises from avoiding double-counting 
of the pairwise gravitational interactions. So while \autoref{eq:grav_energy_conservation_update} was correctly obtained 
by \cite{arepo} (whose work was the primary inspiration for this approach), their derivation that integrates the work 
done over the volume of a zone does not correctly account\footnote{They argued that this volume integral 
of the work can be converted into a surface integral by use of the divergence theorem (see their Equation 93). That 
application of the divergence theorem is mathematically incorrect, because their formulation of the integrand cannot actually 
be expressed only as the divergence of a vector field.} for this behavior. It is only by accounting for the global nature of the
potential that we can obtain the correct factor of $1/2$ and have a correct accounting of the energy within a zone.
The insight of applying the source term on the zone faces to conserve energy is still valid when 
considering energy transfer between the gas energy and a potential field if the change in energy 
is explicitly localizable. Indeed, we employ this formulation in \autoref{sec:rotation} for the case of the rotational potential.

@@

The update to the energy equation can be determined by taking the dot product of the velocity
with the momentum source terms. The Coriolis term vanishes identically, and so
the Coriolis term does no work on the fluid. The update from the centrifugal force becomes
\begin{equation}
  \left.\frac{\partial(\rho E)}{\partial t}\right|_{\text{rot}} = \frac{1}{\Delta V}\int \rho \mathbf{u} \cdot \mathbf{f}^R\, dV,
\end{equation}
with $\mathbf{f}^R \equiv  -{\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right)$. 
This expression is identical in form to the gravity source under the interchange of $\mathbf{g}$ with $\mathbf{f}^R$.
As observed by \cite{marcello:2012}, we can similarly write down a rotational potential,
\begin{equation}
  \Phi^R = -\frac{1}{2} \left| {\bm\omega} \times \mathbf{r} \right|^2.
\end{equation}
In the presence of rotation the total energy becomes:
\begin{equation}
  (\rho E_{\text{tot}}) = (\rho E) + \frac{1}{2} \rho \Phi + \rho \Phi^R.
\end{equation}
Given that we can write down a potential energy for the rotation field, then we can use the machinery of 
\autoref{sec:gravity_hydro_coupling}. We again continue to evolve explicitly an equation for 
the gas energy, and allow it to change in response to work done by or on the rotational potential.
\begin{align}
  \left.\Delta(\rho E)\right|_{\text{rot}} &= - \sum_{f} \Delta \rho_{f} (\Phi^R_{f+1/2} - \Phi^R_{f-1/2})
\end{align}
Again we observe that this results a change in the global integral of $(\rho E)$, but keeps 
the global integral of $(\rho E_{\text{tot}})$ conserved (except at physical domain boundaries).

>>>>

%==========================================================================
% Title
%==========================================================================
\title{White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification}

@@

For the energy equation, the central challenge is to write down a form of the 
discretized energy equation that explicitly conserves total energy when 
coupled to gravity. When gravity is included, the conserved total energy
over the entire domain is
\begin{equation}
  \int \rho E_{\text{tot}}\, dV = \int dV \left(\rho E + \frac{1}{2}\rho\Phi\right), \label{eq:total_energy_gravity}
\end{equation}
where $\rho E$ is the total gas energy from the pure hydrodynamics equation. 
The factor of 1/2 in the gravitational energy is necessary for simulations with
self-gravity to prevent double-counting of interactions (since in dynamical evolution
the relevant gravitational potential energy is $\rho \Phi$ and the gravitational force
is $\rho \mathbf{g}$). Historically many simulation codes with gravity have not used
a conservative formulation of the energy equation, but it is straightforward to do so.
Our approach, and the discussion that follows, is based on that of \cite{arepo}.

Conservation of total energy requires that a change in gravitational energy is compensated
for by a change in gas energy, and that energy changes due to mass transfer are explicitly and 
exactly tracked. Suppose that we have some fluid mass $\Delta M_{i+1/2} = \Delta \rho_{i+1/2} \Delta V$ leave some zone
with index $i$ and enter some zone with index $i+1$. The subscript indicates that the mass change is
occurring at the interface between the two zones, at index $i+1/2$. The work done by the gravitational
force on the gas is $\Delta (\rho E) = W = \int F dx = (\Delta M_{i+1/2}\ g_{i+1/2}) (\Delta x / 2)$,
where $g_{i+1/2}$ is the gravitational acceleration at the interface. The second term in parentheses
is just the distance from the zone center to the zone edge: once the mass leaves the zone edge, it no longer
needs to be tracked. To second order, $g_{i+1/2} = -(\Phi_{i+1} - \Phi_{i}) / \Delta x$, and also to second order the potential
at the interface is given by $\Phi_{i+1/2} = (\Phi_{i+1} + \Phi_i) / 2$, so we can equivalently view the work done
as $W = -\Delta M_{i+1/2} (\Phi_{i+1/2} - \Phi_i)$. Physically, this is just the negative of the gravitational
potential energy change as the fluid is pushed from the cell center potential to the cell edge potential,
exactly as the work-energy theorem implies.

Now, in a hydrodynamics code mass changes correspond to hydrodynamic fluxes. In particular,
the continuity equation tells us that the mass flux $F_\rho = \rho^{n+1/2}_{i+1/2} v^{n+1/2}_{i+1/2}$ yields
an integrated mass motion through the interface $i+1/2$ over a timestep $\Delta t$ of:
\begin{equation}
  \Delta \rho_{i+1/2} = \frac{\Delta t}{\Delta V} \left(\rho^{n+1/2}_{i+1/2} v^{n+1/2}_{i+1/2} dA\right).
\end{equation}
Note that here $v_{i+1/2}$ is the component of the velocity perpendicular to the zone face, whose
area is $dA$.

Finally, then, we write the update in a zone for the total energy that conserves $(\rho E_{\text{tot}})$ as:
\begin{equation}
  \Delta (\rho E) = -\frac{1}{2}\sum_{f} \Delta \rho_{f} (\Phi_{f+1/2} - \Phi_{f-1/2}),\label{eq:grav_energy_conservation_update}
\end{equation}
where the sum is over the cell faces with indices $f$ and the indices $f+1/2$ and $f-1/2$ refer to 
the zone centers immediately to the left and right in the direction perpendicular to the face.
As long as we record the hydrodynamical fluxes through the zone faces after coming out of the hydrodynamics step, 
we are able to conserve the total energy completely (except for any energy loss or gain through 
physical domain boundaries). In order for the method to be second-order accurate in time, 
we need to use a time-centered $\Phi$ (which can be computed by averaging the time-level $n$ and $n+1$ potentials;
we already have the latter because \castro\ re-computes the potential at the new time after the hydrodynamics step,
and we can apply this energy at the end of the timestep). Note that of course the hydrodynamical
flux is already second-order accurate in time. In passing, we observe that the factor of $1/2$ that
appears in \autoref{eq:grav_energy_conservation_update} has nothing to do with the factor of
$1/2$ that appears in the statement of conservation of total energy, \autoref{eq:total_energy_gravity}.
The former comes simply from the fact that the energy change is evaluated using the mass motion through
a distance of half of the zone width. The latter is needed to ensure that these local changes in energy
are not double-counted when doing a global integral, since the gravitational potential is self-generated.
\autoref{eq:grav_energy_conservation_update} applies to any conservative potential $\Phi$, and we will
use this to our advantage for the rotation forces in \autoref{sec:rotation}.

As observed by \cite{arepo}, this is more accurate than the more common (non-conservative) approach
of evaluating the change in gas energy using the work done $(\mathbf{v} \cdot \rho \mathbf{g})$
by the gravitational force at the cell center. Analytically this form expresses the same core idea as
\autoref{eq:grav_energy_conservation_update} via the work-energy theorem, but a major flaw is that
it evaluates the energy change at the cell center when in fact the mass transfer is happening at
the cell edges. This can result in a significant leaking of energy throughout the course of the
evolution, dramatically affecting the course of the evolution. The standard approach is therefore
unacceptable in the case of a problem like white dwarf mergers, and the fix to this energy
leaking---evaluating the energy transfer at the six zone faces instead of the single zone
center---has a very minor cost in terms of code complexity and computational time.

@@

The update to the energy equation can be determined by taking the dot product of the velocity
with the momentum source terms. The Coriolis term vanishes identically, and so
the Coriolis term does no work on the fluid. The update from the centrifugal force becomes
\begin{equation}
  \left.\frac{\partial(\rho E)}{\partial t}\right|_{\text{rot}} = \frac{1}{\Delta V}\int \rho \mathbf{u} \cdot \mathbf{f}^R\, dV,
\end{equation}
with $\mathbf{f}^R \equiv  -{\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right)$. 
This expression is identical in form to the gravity source under the interchange of $\mathbf{g}$ with $\mathbf{f}^R$.
As observed by \cite{marcello:2012}, we can similarly write down a rotational potential,
\begin{equation}
  \Phi^R = \frac{1}{2} \left| {\bm\omega} \times \mathbf{r} \right|^2.
\end{equation}
In the presence of rotation the conserved total energy becomes:
\begin{equation}
  \int dV (\rho E_{\text{tot}}) = \int dV \left( \rho E + \frac{1}{2} \rho \Phi + \rho \Phi^R \right).
\end{equation}
Given that we can write down a potential energy for the rotation field, then we can use the machinery of 
\autoref{sec:gravity_hydro_coupling}. We again continue to evolve explicitly an equation for 
the gas energy, and allow it to change in response to work done by or on the rotational potential.
\begin{align}
  \left.\Delta(\rho E)\right|_{\text{rot}} &= -\frac{1}{2}\sum_{f} \Delta \rho_{f} (\Phi^R_{f+1/2} - \Phi^R_{f-1/2})
\end{align}

### 2015-09-06 papers/wdmerger_II/paper.tex

%==========================================================================
% Title
%==========================================================================
\title{Double White Dwarf Mergers with CASTRO\\ II. Equilibrium Initial Conditions}

\shorttitle{DWD Mergers. II. Equilibrium ICs}
\shortauthors{Max}

\author{TBD}
%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
We describe a method for constructing equilibrium initial conditions for a binary star system,
for a general equation of state. We extend this method to include the effects of adaptive
mesh refinement in the algorithm.

\end{abstract}
\keywords{hydrodynamics - methods: numerical - supernovae: general - white dwarfs}

>>>>

%==========================================================================
% Title
%==========================================================================
\title{White Dwarf Mergers on Adaptive Meshes\\ II. Collisions}

\shorttitle{WD Mergers. II. Collisions}
\shortauthors{Katz et al. (2015)}

\author{TBD}
%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
We consider the collisions of white dwarfs as possible progenitors of Type Ia 
supernovae.

\end{abstract}
\keywords{supernovae: general - white dwarfs}
